Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
paper s5

\subsection{Rankings and orderings: experiments}\label{sec:exper_ro}

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually considers orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$. The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

As a consequence of this important distinction, when evaluating classical
combinatorial problems, we must convert rankings to orderings by inverting the
permutation, as done in Eq.~\ref{eq:lop} and Eq.~\ref{eq:pfsp}, for the LOP and
PFSP, respectively. This might seem an implementation detail but a failure to use the correct interpretation has grave consequences.

During our experiments, we realized that the \emph{swap} distance used by CEGO
assumes rankings, i.e., it actually measures the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP. We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. Of course, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. Therefore, in the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

\manuel{}{The above replaces the text below?}

\ekhine{}{este primer parago es bastante importante para dejarlo aqui. }
In Section~\ref{sec:preliminaries} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion.

In Section~\ref{sec:permutations} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion.

As a summary of Algorithm~\ref{alg:umm}, we model rankings in our UMM model, generate a ranking $\sigma$, then invert this ranking to obtain an ordering in line~\ref{line:inversion} and evaluate the ordering $\sigma^{-1}$ in the objective function. This might seem an implementation detail but for the sake of consistency with the literature, we will expect objective functions to evaluate ordering. A natural question is what happens if we did not invert the ranking and evaluated a ranking with an optimization for orderings. For the anxious reader, we advance that the result is similar to the original version of CEGO.

In CEGO, the objects modeled and evaluated are the same. They claim that they use the swap distance but they are using the \ken  distance, meaning that the object model is a ranking. \ekhine{}{esto es demasido rude?}
On the function evaluation the authors interpert their permutation as an ordering, as usual. 
However, there is no inversion between the objects sampled from the model and the objects evaluated. 
This means that both are rankings (or both are orderings). A natural question is what happens if we did invert the ranking. We have carried out all the experimentation with this modification of CEGO and we have denote this version CEGO\_inv. The original version of CEGO is denoted CEGO\_orig.

@@

We start our analysis with the three version of UMM. We can see a general trend for all the instances: After the 10 random initial permutations, the fitness of the new solutions decreases with a linear trend along iterations. The (close to) linear decrease is given by the linear decrease in the expected distance of the model which is sampled at each step, described in Section~\ref{sec:sampling}. Overall, the improvement does not seem to converge not even for the longer runs of 400 evaluations.
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations.

The red and the gold lines correspond to CEGO, the original and our modified versions

@@

Table~\ref{tab:results} shows results over all instances evaluated (excluding
the synthetic LOP instances). In particular, we show the mean fitness (and
standard deviation) of the best solution found at the end of each run. We also report the mean runtime in hours and, since CEGO sometimes reached the limit of 144 hours, also the number of fitness evaluations reached by CEGO.

In the LOP, differences are mostly in favour of UMM, due to the time limit of
144 hours, which resulted in CEGO typically achieving less than 200
evaluations. In contrast, UMM was always able to reach 400 evaluations in less
than 12.5 hours. If UMM was allowed to reach the 144 hours limit, the differences
would be even stronger.
%
In the PFSP, the differences are minor and/or not statistically
significant, with the qualification that both algorithms appear to be
performing a random search.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $400$ evaluations).
Our experiments presented here show that UMM is able to match the results of
CEGO, a state-of-the-art Bayesian optimizer for expensive black-box
optimization if the computation time of the optimizers is not a concern.

@@

%\MANUEL{UMM is able to model the LOP landscape
%  at least as well as CEGO as shown by the plots where UMM shows a linear
%  decrease in fitness.}  
%  \EKHINE{me parece mas de CEGO q UMM y parece q UMM
%  solo tiene de ventaja ir rápido. Primero destacaría lo bueno de UMM. Creo q
%  es importante destaacar q UMM va MEJOR o IGUAL en TODOS los experimentos y
%  luego que addemás es más rápido} 
%  \MANUEL{Pero UMM es peor en rec13 y nug12, y
%  en LOP es mejor porque CEGO no llega a 400, es decir, porque UMM es más
%  rápido. No podemos saber qué pasaría si CEGO llegara a 400. Podríamos medir
%  lo que pasa en el momento en que matamos a CEGO y dar también ese valor
%  entonces la diferencia en velocidad no importaría. } 
% The time required to train the surrogate-model of \ekhine{esta establecida la complexity de CEGO???}{UMM increases linearly with the permutation length $n$, while in }
% CEGO  it sharply
% increases with $n$, requiring 
% sometimes more than hour per fitness evaluation as shown in the experimental section. 
% Hence, UMM is a computationally
% feasible alternative for relatively large problem sizes.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the results of a Bayesian
optimizer. Moreover, it is overwhelmingly faster than the alternatives for
large problem sizes.  The results presented here show that, despite the
intrinsic difficulty of the expensive black-box combinatorial scenario, there
are still significant advances yet to be made. Thus, we hope that this paper
will motivate further research. % on such problems.
%\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}

An interesting result is that the behavior of UMM (and the same for CEGO) on
the PFSP (and some QAP instances) is close to random. %The same behavior is
%observed for CEGO, which seems to have been never reported until now.
This result may be due to the distance
metric used, but further analysis is needed to corroborate this
conjecture. Here, we have focused on Kendall's-$\tau$ distance, however, we are certain it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.

>>>>

\subsection{Rankings and orderings: experiments}\label{sec:exper_ro}
\ekhine{}{este primer parago es bastante importante para dejarlo aqui. }

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$. %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

As a consequence of this important distinction, when dealing with data in the form of permutations, in particular in 
combinatorial problems, we must convert rankings to orderings by inverting the
permutation, as done in Eq.~\ref{eq:lop} and Eq.~\ref{eq:pfsp}, for the LOP and
PFSP, respectively. This might seem an implementation detail but a failure to use the correct interpretation has grave consequences.

During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP. We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. Of course, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. Therefore, in the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

%\manuel{}{The above replaces the text below?}
%
%In Section~\ref{sec:preliminaries} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion. 
%
%
%In Section~\ref{sec:permutations} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion. 
% 
%As a summary of Algorithm~\ref{alg:umm}, we model rankings in our UMM model, generate a ranking $\sigma$, then invert this ranking to obtain an ordering in line~\ref{line:inversion} and evaluate the ordering $\sigma^{-1}$ in the objective function. This might seem an implementation detail but for the sake of consistency with the literature, we will expect objective functions to evaluate ordering. A natural question is what happens if we did not invert the ranking and evaluated a ranking with an optimization for orderings. For the anxious reader, we advance that the result is similar to the original version of CEGO. 
%
%In CEGO, the objects modeled and evaluated are the same. They claim that they use the swap distance but they are using the \ken  distance, meaning that the object model is a ranking. \ekhine{}{esto es demasido rude?}
%On the function evaluation the authors interpert their permutation as an ordering, as usual. 
%However, there is no inversion between the objects sampled from the model and the objects evaluated. 
%This means that both are rankings (or both are orderings). A natural question is what happens if we did invert the ranking. We have carried out all the experimentation with this modification of CEGO and we have denote this version CEGO\_inv. The original version of CEGO is denoted CEGO\_orig. 
%

@@

\paragraph{Analysis} We start our analysis with the three version of UMM. We can see a general trend for all the instances: After the 10 random initial permutations, the fitness of the new solutions decreases with a linear trend along iterations. The (close to) linear decrease is given by the linear decrease in the expected distance of the model which is sampled at each step, described in Section~\ref{sec:sampling}. Overall, the improvement does not seem to converge not even for the longer runs of 400 evaluations.
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations.

The red and the gold lines correspond to CEGO, the original and our modified versions. For the reason for the abrupt change between both version for every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the original version, there is a global decrease in the fitness function. However, a closer analysis of each iteration does not show a real convergence to any point. However, with the modified version in which rankings are modeled and orderings are evaluated we can appreciate the a decrease in terms of fitness per evaluation. Overall, the modified CEGO has a very abrupt decrease at the first 50 iterations. At one point, the algorithm seems to have converged and very little improvement is don from the 150-th iteration onwards.  The average running time is approximately 22 hours for the 400 function evaluations, which is 300 times slower than UMM for the same number of function evaluations.

The performance of the UMM is clearly better than the performance of the original CEGO for the same problem. For CEGO\_inv on the other hand improves the results of the UMM with the current experimental setting. However, UMM seems to be improving the quality of the solutions until the last iteration while CEGO\_inv seems to be stucked at a local optimum. Further analysis lines will consider a larger budget of fitness evaluation.

Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
The computational complexity of UMM is \ekhine{}{ $O(n \log n)$ }at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.

@@

%Table~\ref{tab:results} shows results over all instances evaluated (excluding
%the synthetic LOP instances). In particular, we show the mean fitness (and
%standard deviation) of the best solution found at the end of each run. We also report the mean runtime in hours and, since CEGO sometimes reached the limit of 144 hours, also the number of fitness evaluations reached by CEGO.

%In the LOP, differences are mostly in favour of UMM, due to the time limit of
%144 hours, which resulted in CEGO typically achieving less than 200
%evaluations. In contrast, UMM was always able to reach 400 evaluations in less
%than 12.5 hours. If UMM was allowed to reach the 144 hours limit, the differences
%would be even stronger.
%%
%In the PFSP, the differences are minor and/or not statistically
%significant, with the qualification that both algorithms appear to be
%performing a random search.

@@

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $100$, $200$ and $400$ evaluations).
Our experiments presented here show that UMM is able to match the results of
the state-of-the-art Bayesian optimizer for expensive black-box
optimization if the computation time of the optimizers is not a concern.

@@

In this paper, we discuss the importance of the representation of permutations in applied mathematics problems. In the context considered in this paper, permutations can bee seen as rankings or orderings of items. In other problems they could be matchings or cycles. 
While certain functions are
this discussion is usually ommited in the literature.
We discuss the different concepts and show, experimentally, the consequences of misunderstanding both.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the state-of-the-art
optimizer. Moreover, it is overwhelmingly faster than the alternatives for
large problem sizes.  The results presented here show that, despite the
intrinsic difficulty of the expensive black-box combinatorial scenario, there
are still significant advances yet to be made. Thus, we hope that this paper
will motivate further research. % on such problems.
%\MANUEL{I'm thinking about finishing here and commenting out the following paragraph}

In this paper, we focus on the UMM under the \ken distance. Therefore, we consider appropiate problems for it, such as the Linear Ordering Problem and the Permutation Flowshop Scheduling Problem, where the pairwise ordering among the items is relevant (as opposed to the actual exat position of each item in the permutation)

However, we are certain it is
possible to extend UMM to other distance metrics, which will allow us to
dynamically select among various distance metrics for an unknown black-box
% \ekhine{}{estoy totalmente en contra de esto :) }
permutation landscape~\citep{ZaeStoBar2014:ppsn}. Also, as shown by our
experiments, the behavior of the parameters ($r_1$ and $r_2$) of UMM is not
always obvious, thus a more detailed analysis would be needed to provide either
generally good static values or an online adaptation approach. Finally, we have
used here three very different combinatorial problems (LOP, PFSP and QAP) as
black-box benchmarks. However, an even more diverse range of problems would be
needed to understand the behavior of UMM on real-world black-box combinatorial
landscapes.
